---
title: "seminrExtras"
author: "Nicholas .P. Danks and Soumya Ray"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{seminrExtras}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(seminr)
library(seminrExtras)

# set the theme for plotting output
seminr_theme_set(seminr_theme_academic())

# Create measurement model ----
corp_rep_mm_ext <- constructs(
  composite("QUAL", multi_items("qual_", 1:8), weights = mode_B),
  composite("PERF", multi_items("perf_", 1:5), weights = mode_B),
  composite("CSOR", multi_items("csor_", 1:5), weights = mode_B),
  composite("ATTR", multi_items("attr_", 1:3), weights = mode_B),
  composite("COMP", multi_items("comp_", 1:3)),
  composite("LIKE", multi_items("like_", 1:3)),
  composite("CUSA", single_item("cusa")),
  composite("CUSL", multi_items("cusl_", 1:3))
)

alt_mm <- constructs(
  composite("QUAL", multi_items("qual_", 1:8), weights = mode_B),
  composite("PERF", multi_items("perf_", 1:5), weights = mode_B),
  composite("CSOR", multi_items("csor_", 1:5), weights = mode_B),
  composite("ATTR", multi_items("attr_", 1:3), weights = mode_B),
  composite("COMP", multi_items("comp_", 1:3)),
  composite("LIKE", multi_items("like_", 1:3)),
  composite("CUSA", single_item("cusa")),
  composite("CUSL", multi_items("cusl_", 1:3))
)

# Create structural model ----
corp_rep_sm_ext <- relationships(
  paths(from = c("QUAL", "PERF", "CSOR", "ATTR"), to = c("COMP", "LIKE")),
  paths(from = c("COMP", "LIKE"), to = c("CUSA", "CUSL")),
  paths(from = c("CUSA"),         to = c("CUSL"))
)
alt_sm <- relationships(
  paths(from = c("QUAL", "PERF", "CSOR", "ATTR"), to = c("COMP", "LIKE")),
  paths(from = c("COMP", "LIKE"), to = c("CUSA")),
  paths(from = c("CUSA"),         to = c("CUSL"))
)


# Estimate the models ----
established_model <- estimate_pls(
  data = corp_rep_data,
  measurement_model = corp_rep_mm_ext,
  structural_model  = corp_rep_sm_ext,
  missing = mean_replacement,
  missing_value = "-99")

competing_model <- estimate_pls(
  data = corp_rep_data,
  measurement_model = alt_mm,
  structural_model  = alt_sm,
  missing = mean_replacement,
  missing_value = "-99")


```

SEMinR (Ray, Danks, & calero-Valdez, 2026) is a domain specific language for modeling and estimating structural equation models. This is a supplementary package for SEMinR and not a standalone package. 
This package serves to provide additional extra methods and functions that can be used to analyze PLS-SEM models. 

SEMinRExtras provides advanced SEM tools which are compatible with SEMinR. It implements several methods for evaluating PLS-SEM models, including CVPAT (Liengaard et al., 2021; Sharma et al., 2022), Composite Overfit Analysis (COA), Necessary Condition Analysis (NCA; Dul, 2016; Richter et al., 2020), NCA-ESSE (Becker et al., 2026), and congruence testing.

SEMinRExtras also serves to host the example models used in the PLS-SEM in R workbook (Hair et al., 2026).

### Functions

| Function | Description |
|---|---|
| `assess_cvpat()` | CVPAT against LM and IA benchmarks |
| `assess_cvpat_compare()` | Compare predictive loss of two PLS models |
| `assess_coa()` | Composite Overfit Analysis (full pipeline) |
| `predictive_deviance()` | Compute predictive deviance scores |
| `deviance_tree()` | Identify deviant case groups via decision tree |
| `unstable_params()` | Parameter instability analysis |
| `assess_nca()` | Necessary Condition Analysis for PLS-SEM |
| `assess_nca_esse()` | NCA with Effect Size Sensitivity Extension |
| `congruence_test()` | Bootstrapped congruence coefficient testing |

## The demo files for Hair et al. (2026)  

In order to access the demo files for the textbook, you can run the ```demo()``` function after loading the SEMinRExtras package. 

- seminr-help-debugging
- seminr-pls-cvpat
- seminr-pls-nca
- seminr-primer-v2-chap2
- seminr-primer-v2-chap3
- seminr-primer-v2-chap4
- seminr-primer-v2-chap5
- seminr-primer-v2-chap6 
- seminr-primer-v2-chap7 
- seminr-primer-v2-chap8

E.g. ```demo("seminr-pls-cvpat", package = "seminrExtras")```

## The Example model: Corporate Reputation

We are applying the CVPAT process to the corporate reputation example bundled with SEMinR. Since we will be comparing two models, we will first estimate and plot both models. Below you will find the competing and established models which will be compared. 

```{r, fig.show='hold'}
plot(established_model)
plot(competing_model)
```

## Example
```{r, eval=FALSE, echo=TRUE}

# Create measurement model ----
corp_rep_mm_ext <- constructs(
  composite("QUAL", multi_items("qual_", 1:8), weights = mode_B),
  composite("PERF", multi_items("perf_", 1:5), weights = mode_B),
  composite("CSOR", multi_items("csor_", 1:5), weights = mode_B),
  composite("ATTR", multi_items("attr_", 1:3), weights = mode_B),
  composite("COMP", multi_items("comp_", 1:3)),
  composite("LIKE", multi_items("like_", 1:3)),
  composite("CUSA", single_item("cusa")),
  composite("CUSL", multi_items("cusl_", 1:3))
)

alt_mm <- constructs(
  composite("QUAL", multi_items("qual_", 1:8), weights = mode_B),
  composite("PERF", multi_items("perf_", 1:5), weights = mode_B),
  composite("CSOR", multi_items("csor_", 1:5), weights = mode_B),
  composite("ATTR", multi_items("attr_", 1:3), weights = mode_B),
  composite("COMP", multi_items("comp_", 1:3)),
  composite("LIKE", multi_items("like_", 1:3)),
  composite("CUSA", single_item("cusa")),
  composite("CUSL", multi_items("cusl_", 1:3))
)

# Create structural model ----
corp_rep_sm_ext <- relationships(
  paths(from = c("QUAL", "PERF", "CSOR", "ATTR"), to = c("COMP", "LIKE")),
  paths(from = c("COMP", "LIKE"), to = c("CUSA", "CUSL")),
  paths(from = c("CUSA"),         to = c("CUSL"))
)
alt_sm <- relationships(
  paths(from = c("QUAL", "PERF", "CSOR", "ATTR"), to = c("COMP", "LIKE")),
  paths(from = c("COMP", "LIKE"), to = c("CUSA")),
  paths(from = c("CUSA"),         to = c("CUSL"))
)


# Estimate the models ----
established_model <- estimate_pls(
  data = corp_rep_data,
  measurement_model = corp_rep_mm_ext,
  structural_model  = corp_rep_sm_ext,
  missing = mean_replacement,
  missing_value = "-99")

competing_model <- estimate_pls(
  data = corp_rep_data,
  measurement_model = alt_mm,
  structural_model  = alt_sm,
  missing = mean_replacement,
  missing_value = "-99")

# Function to compare the Loss of two models
compare_results <- assess_cvpat_compare(established_model = established_model,
                                        alternative_model = competing_model,
                                        testtype = "two.sided",
                                        nboot = 2000,
                                        technique = predict_DA,
                                        seed = 123,
                                        noFolds = 10,
                                        reps = 10,
                                        cores = 1)


print(compare_results,
      digits = 3)

# Assess the base model ----
assess_results <- assess_cvpat(established_model,
                               seed = 123, 
                               cores = 1)
print(assess_results$CVPAT_compare_LM,
      digits = 3)
print(assess_results$CVPAT_compare_IA,
      digits = 3)

```

## First, conduct CVPAT analysis of the established model.

```{r, echo=TRUE}
# Assess the base model ----
assess_results <- assess_cvpat(established_model,
                               seed = 123, 
                               cores = 1)
print(assess_results$CVPAT_compare_LM,
      digits = 3)
print(assess_results$CVPAT_compare_IA,
      digits = 3)
```

The established model has significantly lower predictive loss compared to both the naive benchmark IA and the LM model. Thus, we can say that the established model has predictive relevance.   

Now we compare the results:

```{r, echo=TRUE}
# Function to compare the Loss of two models
compare_results <- assess_cvpat_compare(established_model = established_model,
                                        alternative_model = competing_model,
                                        testtype = "two.sided",
                                        nboot = 2000,
                                        technique = predict_DA,
                                        seed = 123,
                                        noFolds = 10,
                                        reps = 10,
                                        cores = 1)

print(compare_results, 
      cores = 1,
      digits = 3)
```

The established model has significantly lower predictive loss compared to the competing model. Thus, we can say that the established model has superior predictive performance compared to the competing model.

## Necessary Condition Analysis (NCA)

NCA (Dul, 2016) tests whether a predictor is a *necessary condition* for an outcome -- i.e., whether a certain level of X is required (but not sufficient) for a certain level of Y. This complements the sufficiency-based logic of PLS-SEM regression.

`assess_nca()` auto-detects direct predictors from the structural model and runs ceiling-based NCA analysis using internal CE-FDH and CR-FDH algorithms.

```{r, eval=FALSE, echo=TRUE}
library(seminr)
library(seminrExtras)

# Estimate a simple model
mobi_mm <- constructs(
  composite("Image",        multi_items("IMAG", 1:5)),
  composite("Value",        multi_items("PERV", 1:2)),
  composite("Satisfaction", multi_items("CUSA", 1:3)),
  composite("Loyalty",      multi_items("CUSL", 1:3))
)

mobi_sm <- relationships(
  paths(from = c("Image", "Value"), to = "Satisfaction"),
  paths(from = "Satisfaction", to = "Loyalty")
)

mobi_pls <- estimate_pls(data = mobi,
                          measurement_model = mobi_mm,
                          structural_model  = mobi_sm)

# Run NCA -- predictors auto-detected from structural model
nca_result <- assess_nca(mobi_pls,
                          target = "Satisfaction",
                          test.rep = 1000,
                          seed = 123)

# Effect sizes and significance
print(nca_result)

# Full summary with bottleneck tables
summary(nca_result)

# Visualize
plot(nca_result, type = "effects")   # Effect size bar chart
plot(nca_result, type = "scatter")   # Ceiling line scatter plots
```

A predictor is identified as a necessary condition when it has an effect size d >= 0.1 and a significant permutation test (p < 0.05), following guidelines from Richter et al. (2020).

## NCA-ESSE: Effect Size Sensitivity Extension

NCA-ESSE (Becker et al., 2026) extends standard NCA by assessing how sensitive effect sizes are to extreme response patterns. It systematically varies the ECDF ceiling threshold and compares empirical effect size changes against a theoretical benchmark derived from a joint uniform distribution.

```{r, eval=FALSE, echo=TRUE}
# Run NCA-ESSE on the same model
esse_result <- assess_nca_esse(mobi_pls,
                                target = "Satisfaction",
                                thresholds = seq(0, 0.05, by = 0.005),
                                seed = 123)

# View results
print(esse_result)

# Summary tables (Table A2 style)
summary(esse_result)

# Sensitivity plot (Fig. 4 in Becker et al., 2026)
plot(esse_result, type = "sensitivity")

# Difference plot (Fig. 6 in Becker et al., 2026)
plot(esse_result, type = "difference")
```

At each threshold `t`, observations whose joint ECDF_NCA value P(X <= x, Y >= y) is at most `t` are treated as extreme upper-left cases and excluded. The uniform benchmark d = t(1 - ln(t)) gives the expected effect size increase if no true necessity exists. Empirical effect sizes that consistently exceed the benchmark provide evidence for a robust necessary condition.

## Composite Overfit Analysis (COA)

COA detects observation-level overfitting in PLS composite models. It computes predictive deviance (the gap between in-sample fitted and out-of-sample predicted scores), then uses a decision tree to identify subgroups where the model overfits.

```{r, eval=FALSE, echo=TRUE}
library(seminr)
library(seminrExtras)

# Estimate a model
corp_mm <- constructs(
  composite("COMP", multi_items("comp_", 1:3)),
  composite("LIKE", multi_items("like_", 1:3)),
  composite("CUSA", single_item("cusa")),
  composite("CUSL", multi_items("cusl_", 1:3))
)

corp_sm <- relationships(
  paths(from = c("COMP", "LIKE"), to = "CUSA"),
  paths(from = "CUSA", to = "CUSL")
)

corp_model <- estimate_pls(
  data = corp_rep_data,
  measurement_model = corp_mm,
  structural_model  = corp_sm,
  missing = mean_replacement,
  missing_value = "-99")

# Run full COA pipeline
coa_result <- assess_coa(corp_model,
                          focal_construct = "CUSL",
                          noFolds = 10, reps = 1, cores = 1,
                          seed = 123)

# Print results
print(coa_result)
summary(coa_result)

# Visualize
plot(coa_result, type = "pd")      # Predictive deviance distribution
plot(coa_result, type = "tree")    # Decision tree
plot(coa_result, type = "groups")  # Deviant group highlights
```

COA proceeds in four steps: (1) compute predictive deviance via cross-validated prediction, (2) grow a decision tree to identify deviant subgroups, (3) analyze parameter instability by re-estimating the model without each deviant group, and (4) inspect group rules and competing splits.

# References
- Becker, J.-M., Richter, N. F., Ringle, C. M., & Sarstedt, M. (2026). Must-have, or maybe not? A sensitivity-based extension to necessary condition analysis. Journal of Business Research, 206, 115920.
- Dul, J. (2016). Necessary Condition Analysis (NCA): Logic and Methodology of 'Necessary but Not Sufficient' Causality. Organizational Research Methods, 19(1), 10-52.
- Hair, J.F. (Jr), Hult, T.M., Ringle, C.M., Sarstedt, M., Danks, N.P., and Adler, S. (2026). Partial Least Squares Structural Equation Modeling (PLS-SEM) Using R (Second Edition) - A Workbook. Springer.
- Liengaard, B. D., Sharma, P. N., Hult, G. T. M., Jensen, M. B., Sarstedt, M., Hair, J. F., & Ringle, C. M. (2021). Prediction: coveted,yet forsaken? Introducing a cross‐validated predictive ability test in partial least squares path modeling. Decision Sciences, 52(2), 362-392.
- Ray, S., Danks, N.P., Calero Valdez, A. (2026) SEMinR: Domain-specific language for building, estimating, and visualizing structural equation models in R.
- Richter, N. F., Schubring, S., Hauff, S., Ringle, C. M., & Sarstedt, M. (2020). When predictors of outcomes are necessary: guidelines for the combined use of PLS-SEM and NCA. Industrial Management & Data Systems, 120(12), 2243-2267.
- Sharma, P. N., Liengaard, B. D., Hair, J. F., Sarstedt, M., & Ringle, C. M. (2022). Predictive model assessment and selection in composite-based modeling using PLS-SEM: extensions and guidelines for using CVPAT. European journal of marketing, 57(6), 1662-1677.
